\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{enumerate}

\pagestyle{empty}

\begin{document}

\begin{center}
  \textbf{\large MATH 3305 (Spring 26) \hfill Instructor: Zhao}\\[4pt]
  \textbf{\large Homework \#6 \hfill Due: Tue 3/10 (Assigned 3/3)}
\end{center}

\bigskip

\begin{enumerate}
  \item Page 70, Exercise 2.2.5 (page 186 of library e-book). PLUS (d) Graph $F(x)$.

    An automobile dealer has found that $X$, the number of cars customers buy each week, follows the probability distribution
    \[
      f(x) = \begin{cases} \dfrac{kx^2}{x!}, & x = 1, 2, 3, 4. \\[6pt] 0, & \text{otherwise.} \end{cases}
    \]
    \begin{enumerate}[(a)]
      \item Find $k$.
      \item Find the probability the dealer sells at least two cars in a week.
      \item Find $F(x)$, the cumulative distribution function.
      \item Graph $F(x)$.
    \end{enumerate}

    Note: you may use R function \texttt{plot(stepfun(x, F, right=T), ylab="$F(x)$", main="your title")}, where \texttt{x} and \texttt{F} are two vectors of values. For example, $x = c(1,2,5)$, $F = c(0, 0.3, 0.5, 1.0)$.

  \item Page 71, Exercise 2.2.13 (page 187 of library e-book).

    A random variable, $X$, has the distribution function
    \[
      F(x) = \begin{cases} 0, & x < -1 \\[4pt] \dfrac{1}{3}, & -1 \le x < 0 \\[6pt] \dfrac{5}{6}, & 0 \le x < 2 \\[6pt] 1, & x \ge 2 \end{cases}
    \]
    Find the probability distribution function, $f(x)$.

  \item A bag contains eight numbered pieces of paper using numbers 1, 2, 3, 4, 5, 6, 7, and 8. Let $X$ be the number on a randomly selected piece of paper.
    \begin{enumerate}[(a)]
      \item Find the probability mass function, $f(x)$.
      \item Find $\mathbb{E}(X)$, the mean of $X$.
    \end{enumerate}

  \item A term life insurance policy will pay a beneficiary a certain sum of money upon the death of the policyholder. These policies have premiums that must be paid annually. Suppose a life insurance company sells a \$200,000 one-year term life insurance policy to a 49-year-old female for \$600. According to the \textit{National Vital Statistics Report}, Vol.~47, No.~28, the probability the female will survive the year is 0.9975.
    \begin{enumerate}[(a)]
      \item Compute the expected profit to the insurance company for selling such a policy.
      \item Interpret the value of expected profit in (a) in the context of the problem.
    \end{enumerate}

  \item \textit{Do all random variables have a mean?} Let $X$ be a discrete random variable having the probability mass function
    \[
      f(x) = \frac{6}{\pi^2 x^2}, \quad x = 1, 2, 3, \cdots
    \]
    Show that $\mathbb{E}(X)$ does not exist (i.e., it is not finite). \hfill \textcolor{blue}{Hint: Harmonic series diverges.}
\end{enumerate}

\begin{center}
  \textbf{\large Show work!}
\end{center}

\end{document}
