\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{enumerate}

\pagestyle{empty}

\begin{document}

\begin{center}
  \textbf{\large MATH 3305 (Spring 26) \hfill Instructor: Zhao}\\[4pt]
  \textbf{\large Homework \#7 \hfill Due: Mon 3/23 (Assigned 3/13)}
\end{center}

\bigskip

\begin{enumerate}
  \item Page 79, Exercise 2.3.3 (page 201 of library e-book) with 0.1 added to each of the two probabilities.

    A small manufacturing firm sells 1 machine per month with probability 0.3; it sells 2 machines per month with probability 0.1; it never sells more than 2 machines per month. If $X$ represents the number of machines sold per month,

    \textbf{With 0.1 added to each of the two probabilities:} $P(X=1) = 0.4$, $P(X=2) = 0.2$.
    \begin{enumerate}[(a)]
      \item Find the mean and variance of $X$.
      \item If the monthly profit is $2X^2 + 3X + 1$ (in thousands of dollars), find the expected monthly profit.
    \end{enumerate}

  \item Page 79, Exercise 2.3.8 (page 203 of library e-book).

    A coin, loaded to show heads with probability $2/3$, is tossed until heads appear or until 5 tosses have been made. Let $X$ denote the number of tosses made. Find the mean and variance of $X$.

  \item Given $\mathbb{E}(2X + 5) = 10$ and $\mathbb{E}\!\left((2X + 5)^2\right) = 125$. Determine
    \begin{enumerate}[(a)]
      \item The mean of $X$, $\mu_X$.
      \item The variance of $2X + 5$, $\sigma_{2X+5}^2$.
      \item The variance and standard deviation of $X$, $\sigma_X^2$ and $\sigma_X$.
    \end{enumerate}

  \item Label the sides of a fair 4-sided die (tetrahedron) with numbers 1, 2, 3, and 4. Roll the die and let $X$ be the number facing down.
    \begin{enumerate}[(a)]
      \item Find the mean and variance of $X$.
      \item What is variance of $3X - 2$?
    \end{enumerate}

  \item In a small voting precinct, 160 voters favor a candidate and 40 voters oppose the candidate. A sample of 3 voters is selected at random.
    \begin{enumerate}[(a)]
      \item What is the exact probability that all voters in the sample will oppose the candidate?
      \item Exactly what underlying distribution is used in (a)? Find the mean and variance of the distribution.
    \end{enumerate}

  \item A class of 35 students consists of 20 females and 15 males. Select one student at random from the class. Let $X = 1$ if a female student is selected and $X = 0$ otherwise.
    \begin{enumerate}[(a)]
      \item Determine the distribution of $X$ including name of the distribution and value(s) of parameter(s).
      \item Find the mean, and variance of $X$.
    \end{enumerate}
\end{enumerate}

\begin{center}
  \textbf{\large Show work!}
\end{center}

\end{document}
