\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{enumerate}
\usepackage{xcolor}

\pagestyle{empty}

\begin{document}

\begin{center}
  \textbf{\large MATH 3305 (Spring 26) \hfill Instructor: Zhao}\\[4pt]
  \textbf{\large Homework \#8 \hfill Due: Tue 3/31 (Assigned 3/25)}
\end{center}

\bigskip

\begin{enumerate}
  \item Page 87, Exercise 2.5.12ab (page 218 of library e-book). PLUS

    (c) Find the mean and standard deviation for $X$.

    The probability of being able to log on a computer system from a remote terminal during a busy period is 0.7. Suppose that 10 independent attempts are made and that $X$ denotes the number of successful attempts.
    \begin{enumerate}[(a)]
      \item Write an expression for the probability distribution function, $f(x)$.
      \item Find $P(X \ge 5)$.
      \item Find the mean and standard deviation for $X$.
    \end{enumerate}

  \item Page 87, Exercise 2.5.17a (page 220 of library e-book) with 90 changed to \textcolor{blue}{93}.

    An airline knows that 10\% of the people holding reservations on a given flight will not appear. The plane holds \textcolor{blue}{93} people. If 95 reservations have been sold, find the probability that the airline will be able to accommodate everyone appearing for the flight.

  \item In a small voting precinct, 160 voters favor a candidate and 40 voters oppose the candidate. A sample of 3 voters is selected at random.
    \begin{enumerate}[(a)]
      \item Use a binomial probability to approximate the probability that all three voters selected oppose the candidate.
      \item Is the criterion for a good approximation met? Is the approximated probability in (a) close to 0.0075, the true probability from Problem 5(a) of HW7?
    \end{enumerate}

  \item Label the sides of a fair 4-sided die (tetrahedron) with numbers 1, 2, 3, and 4. Roll the die and let $X$ be the number facing down. Determine the MGF of $X$.
\end{enumerate}

\begin{center}
  \textbf{\large Show work!}
\end{center}

\end{document}
